Title

### 2018-09-20 thesis/paper/paper.tex
Changed emphasis to sc

\chapter{Preliminaries}\label{cha_preliminaries}
\section{Wikidata}\label{sec_wikidata}
Wikidata is the knowledge base sister project of Wikipedia. It is a public, open database system that, at its core, stores statements about specific items. The basic unit of information is a triple subject-predicate-object stating an item (subject) has a property (predicate) which is either another item or a value of some kind (object). Triples store the relations of entities with each other or with specific data. An entity is either an item, which represents a topic, concept or real world entity, or a property, which denotes a relation an item has to another or with a value. They have numeric identifiers that are prefixed with Q and P and will be referred to as Q- and P-IDs, respectively. In the future there will be a third entity type, lexeme. This is already incorporated into some constraints, but as of September 2018 they are not deployed. A value is one of several data types shown in Table \ref{tab_datatypes}, although it is also possible to use \emph{none} or \emph{some} as values for statements that have no value (akin to negation) or have a currently unknown value. Every property determines the data type it accepts as an object. \\

@@

While this is similar to RDF and Wikidata has in fact been exported as RDF \cite{EGKMV2014}, the data structure is more complex than simple triples. It allows to store additional information about statements. This is done via a statement ID, which is assigned to every triple storing information about an item. Statements have one of three ranks, \emph{normal}, \emph{preferred}, and \emph{deprecated} which are used to differentiate similar statements on the same item. For example, the \emph{preferred} rank would be given to the current \emph{head of state} (P35) of a country, while all statements denoting former \emph{heads of state} have the rank \emph{normal}. \emph{Deprecated} is used on statements that where once considered correct but are not anymore, e.g., \emph{Chelsea Manning} (Q298423) having the \emph{sex or gender} (P21) \emph{male} (Q6581097).

Every statement can have added qualifiers, which consist of the corresponding statement id, a predicate, and an object. Qualifiers provide contextual information for statements. The most common context is the time frame in which the statement was accurate, e.g., stating the \emph{start time} (P580) and \emph{end time} (P582) of the term of a head of state, but other information is possible, e.g. determination method for measurements or the elevation above sea level for the highest point. Additionally, references can be added to statements.

@@

\begin{example}
Figure \ref{fig_example} is an excerpt from the entity page of the clipper \emph{Cutty Sark} (Q171255) \footnote{\url{https://www.wikidata.org/wiki/Q171255}}. An entity page lists all statements with the entity as the subject. It then groups all statements by their property, in this case \emph{length} (P2043). For the length of the Cutty Sark there is only one value, 85.4 metres, but this could include multiple values, e.g., if there were different ways to determine the length. This statement is further qualified with \emph{criterion used} (P1013) as the qualifier property, which specifies how the value was determined. In this case the qualifier value \emph{length overall} (Q2358152) was taken. The statement is backed up by one reference citing the property \emph{reference URL} (P854) pointing to the reference value, which is an URL confirming the length of 85.4 metres.

@@

\section{Datalog}
Datalog is a logic query language. It is fully declarative and uses rules to derive additional knowledge from given facts. The most basic domain element in datalog is a \emph{constant}, which will be denoted as \emph{someConstant}. In the context of Wikidata, any item or property, e.g. \emph{Earth} (Q2) or \emph{head of government} (P6), could be represented as a constant. Relations in datalog are expressed using \emph{predicates} with a specified arity \(n\), written as predicate\([n]\). To stay close to Wikidata in our examples, let us take the predicate triple\([3]\) as an example. This would be used to express the \emph{item}-\emph{property}-\emph{value} relation with each position representing one part of the triple.\\
Additionally, datalog also provides \emph{variables}, which will be capitalized, like \(X\). The sets of predicates {\bf P}, constants {\bf C} and variables {\bf V} are mutually disjoint. Constants and variables are collectively known as \emph{terms} and are, together with predicates, used to form \emph{atoms}. An atom has the form predicate\((t_1, \ldots , t_n)\) with \(n\) as the arity of the predicate and \(t_1,\ldots,t_n\) terms.\\
Atoms are the building blocks of \emph{rules}, which are used to derive information in datalog. A rule is a premise that implies a conclusion, and both are conjunctions of atoms.
Let \(P_n\) denote  premise atoms and \(C_n\) the conclusion atoms. The variables used in one rule are divided into two mutually disjoint sets \(V_p\) and \(V_r\) which are the variables occurring only in its premise atoms and occurring in conclusion atoms and premise atoms, respectively. The general from of a rule is (with \(k \in \mathbb{N}^+\), \(l \in \mathbb{N}\))
\begin{equation}
\forall V_p \forall V_r.(C_1 \wedge \ldots \wedge C_k \leftarrow P_1 \wedge \ldots \wedge P_l)
\end{equation}

If \(V_c = \emptyset\) the rule is called a \emph{datalog rule} and if \(l = 0\) it is called a \emph{fact}. If a fact contains no variables but only constants, it is a \emph{ground rule}. Since all \(V_p\) and \(V_r\) are  quantified, the preceding universal quantifiers \(\forall V_p \forall V_r\) will be omitted for the rules presented in this work.

Within the rule predicates we can differentiate between two different schemas: the \emph{extensional schema} and the \emph{intensional schema}, called EDB and IDB (extensional/intensional database). An EDB predicate may only occur in the premise of a rule, meaning it can never be derived. All predicates that appear in the conclusion at least once are IDB predicates.

\begin{example}
For the following example, taken from Wikidata, descriptive names will be used instead of the IDs to increase readability. The individual entities and properties are introduced with their ID beforehand, so the example can be retraced. The species \emph{blue whale} (Q42196) has the \emph{parent taxon} (P171) \emph{balaenoptera} (Q133320), which in turn has the parent taxon \emph{rorqual} (Q232829). The predicates used are triple\_EDB\([3]\) as the input and isInTaxon\_IDB\([2]\), which will be derived. The basic facts are:
\[\text{triple\_EDB}(\text{blue whale}, \text{parent taxon}, \text{balaenoptera})\]
\[\text{triple\_EDB}(\text{balaenoptera}, \text{parent taxon}, \text{rorqual})\]
From this we can draw conclusions. The first step is to represent the already existing facts as IDB:
\[\text{isInTaxon\_IDB}(X, Y) \leftarrow \text{triple\_EDB}(X, \text{parent taxon}, Y)\]
Finally the transitive closure of this relation can be derived with
\[\text{isInTaxon\_IDB}(X, Y) \leftarrow \text{isInTaxon\_IDB}(X, Z) \wedge \text{isInTaxon\_IDB}(Z, Y)\]
\end{example}

@@

To apply a constraint to a property, a statement <property> <P2302> <constraint> is added to the properties item page, where <property> is the property that should be constrained, <P2302> (property constraint) denotes that the property has a constraint and <constraint> is the specific constraint. Note that a constrained statement means a statement whose predicate is a property that has a specific constraint. A simple example for this would be \emph{distinct values} (Q21502410), stating that values of the property must be unique across all constrained statements. For this case no additional information regarding the constraint is necessary and violation could be found by surveying all constrained statements and comparing their values.\\
Constraints can be further specified by using qualifiers on the constraint statement. These qualifiers are either optional or mandatory, and may be applied once or multiple times. For example, the \emph{none of} constraint (Q52558054) states that values for the specific property may not be in a given set of disallowed values. This set is specified using qualifiers with property \emph{item of} (P2305), which list all the disallowed values. At least one such qualifier must be present for the statement.\\
More complex dependencies are also possible. A constraint can depend on all constrained statements of a specific item (\emph{single value} constraint [Q19474404]), on all statements of the item specified as the value of a constrained statement (\emph{value requires statement} constraint [Q21510864]), or on all statements using the subclass of-predicate (\emph{type} constraint [Q21503250]).

@@

\section{Rules}
Table \ref{tab_property_constraints} shows that most constraints cannot be modelled without at least some amount of preprocessing. Only \emph{scope}, \emph{conflicts with}, \emph{allowed entity types} and \emph{none of} can be modelled directly. This is due to the fact that \emph{conflicts with} and \emph{none of} exclude a limited and specified set of statements from validity and these cases can thus be directly taken as violations. \emph{scope} and \emph{allowed entity types} work very similar, since there are only three respectively two possible values and violations are simply the options that have not been allowed. The rules for these constraints are described in Section \ref{sec_simple_constraints}.\\

A problem that occurs multiple times is the need to determine inequality between constants. \emph{distinct value}, \emph{allowed units}, \emph{allowed qualifiers}, \emph{one of}, \emph{single best value} and \emph{one-of qualifier value} each require a rule to determine if a value, unit, or property is not equal to another. To this end a certain amount of preprocessing is necessary, most notably the extraction of all constant pairs whose inequality has to be determined for finding the violations. To avoid putting and undue amount of processing outside of datalog, this extraction will not check more than that the constant is in the right position relative to the constrained statement, e.g., all values on statements constrained by \emph{distinct value} or all qualifiers on statements constrained by \emph{allowed qualifiers}. From this set inequality will be generated, different versions of which will be introduced in Section \ref{sec_inequality} and evaluated with regards to their performance in Section \ref{sec_constraints_using_inequality}.\\

Negation as a general problem cannot be solved in datalog. One could pre-process every necessary fact to be negated and add this to the ground facts, but this is not the purpose of this work. However, not all constraints require a general negation. \emph{item requires statement}, \emph{value requires statement}, \emph{inverse }and \emph{symmetric} are all violated if a statement does not exist on a specific item. This problem can be solved with limited amounts of preprocessing by leveraging a preprocessed statement order per item and 'iterating' through the statements until the required conditions are met.\\
A similar approach can be taken to find items with exactly one statement matching certain criteria, required for 	\emph{multi-value}. The idea is to ask the question if a second statement exists, then evaluating it as above. Both methods will be explained in Section \ref{sec_constraints_using_statement_non}.

\subsection{Representation of Wikidata as Facts}\label{subsec_representation}
To represent the content of Wikidata as facts, the following predicates are introduced:
\begin{table}[H]
\caption{EDB predicates}\label{tab_predicates}
\begin{tabular}{llp{8.5cm}}
Predicate & Positions & Description \\
\hline
statement\_EDB & \makecell{statementID \\ entityID \\ propertyID \\ value (entityID or string)} & represents basic statements \\
\hline
qualifier\_EDB & \makecell{statementID \\ propertyID \\ value} & represents qualifiers on statements \\
\hline
reference\_EDB & \makecell{statementID \\ reference hash \\ propertyID \\ value} & represents references on statements; reference hash groups the property-value pairs belonging to one reference \\
\hline
item\_EDB & entityID & marks an entity as an item \\
property\_EDB & entityID & marks an entity as a property \\
\hline
unit\_EDB & \makecell{value \\ unitID} & denotes which unit a value has \\
\hline
rank\_EDB & \makecell{statementID \\ rank} & denotes the rank a statement has
\end{tabular}
\end{table}

@@

\subsection{Task}\label{subsec_task}
Each property constraint will be modelled and evaluated as one set of rules per property per constraint, meaning each violating statement will be of a form similar to
\begin{align*}
&\text{violation\_statement}(S, E, \text{propertyConstant}, V) &\leftarrow \text{atoms determining the violation} \\
&\text{violation\_qualifier}(S, \text{propertyConstant}, V) &\leftarrow \text{atoms determining the violation} \\
&\text{violation\_reference}(S, H, \text{propertyConstant}, V) &\leftarrow \text{atoms determining the violation}
\end{align*}
where propertyConstant denotes the constrained property. The variables \(S\), \(I\), \(V\), and \(H\) are used for statement ID, entity ID, value and reference hash, respectively.

The task for each constraint is thus: Given the constraint and a constrained property, potentially with qualifiers, which rule finds all violations of the rule in a given set of triples, qualifiers and references if applicable? The given rule set should return all violations if queried with the following atoms:
\begin{align*}
&\text{violation\_statement}(S, E, P, V) \\
&\text{violation\_qualifier}(S, P, V) \\
&\text{violation\_reference}(S, H, P, V)
\end{align*}

@@

{\bf Constraint:} This subsection starts with a table showing basic properties of the constraint. These include the required data, relevant qualifiers, and possible violations of the constraint. The required data states which part of the data in relation to the constrained statement are restricted or need to exist. The qualifiers list their cardinality, meaning if they are required and if they may be used multiple times, and a short description. The violations list the conflicting (part of a) statement in relation to the constrained statement. The first constraint, \emph{property scope}, will be used as an example to explain the table in detail.

@@

\subsection{Property scope (Q53869507)}
\paragraph{Constraint}
The \emph{property scope} constraint specifies if a property may be used as predicate in a statement, qualifier, or reference.
\begin{table}[H]
\caption{property scope constraint (Q53869507)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Required data & constrained statement, qualifier and reference \\
\hline
Qualifiers & \makecell{\emph{property scope} (P5314) -- 1..3 \\ must be \emph{as main value} (Q54828448), \emph{as qualifier} (Q54828449), \\ or \emph{as reference} (Q54828450) \\ specifies allowed use in statements, qualifiers, or references} \\
\hline
Violation & constrained property in a position that is not allowed \\
\hline
\end{tabularx}
\end{table}

The required data specifies which data, in relation to the constrained statement, needs to be evaluated to conclude if the constraint was violated. The used abbreviations are listed in Table \ref{tab_abbreviations}. In this case, this is simply each statement, qualifier, or reference whose predicate is the constrained property.\\
The \emph{property scope} constraint requires the scope, meaning statement, qualifier, or reference, to be set via qualifier \emph{property scope} (P5314). The cardinality is 1..3 because it is mandatory and has only three possible values.\\
It is violated if a statement, qualifier, or reference has a predicate with the \emph{property scope} constraint that does not allow this scope.\\

\begin{example}
The \emph{reference URL} (P854) property is restricted to the \emph{property scope} (P5314) \emph{as reference} (Q54828450), because it is to be used to mark a URL that substantiates a statement.
\end{example}

\paragraph{Rules}
The \emph{property scope} constraint requires no context information to be violated; the triple, qualifier, or reference with the constrained property is enough. This means violations can be found using these rules:
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow \\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V)\\
\end{split}\\
\begin{split}
\text{violation\_qualifier}&(S, \text{propertyConstant}, V) \leftarrow \\
&\text{qualifier\_EDB}(S, \text{propertyConstant}, V)\\
\end{split}\\
\begin{split}
\text{violation\_reference}&(S, \text{propertyConstant}, V) \leftarrow \\
&\text{reference\_EDB}(S, H, \text{propertyConstant}, V)\\
\end{split}
\end{align}
Which of these rules need to be used depends on the values of the \emph{property scope} qualifier. If a concrete constrained on a property does not have \emph{as main value} as a value of the qualifier, the first rule is used. The case for \emph{as qualifier} and \emph{as reference} is analogous with the second and third rule, respectively.

\subsection{Conflicts with (Q21502838)}
\paragraph{Constraint}
The \emph{conflicts with} constraint forbids statements on constrained entities whose predicate is a specified \emph{conflicting property}. If \emph{conflicting values} have been specified as well, the constraint is weakened and only violated by statements if in addition to the predicate being a \emph{conflicting property}, the object is one of the \emph{conflicting values}.
\begin{table}[H]
\caption{Conflicts-with constraint (Q21502838)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Required data & entity statements \\
\hline
Qualifiers & \makecell{\emph{conflicting property} (P2306) -- 1 \\ property that may not be predicate of entity statements \\
\emph{conflicting values} (P2305) -- 0..* \\ values that may not be objects of entity statements with the \emph{conflicting property}} \\
\hline
Violation & \makecell{entity statement with the conflicting property \\ entity statement with the conflicting property and one of the conflicting values} \\
\hline
\end{tabularx}
\end{table}

\begin{example}
The \emph{author} (P50) constraint has (among others) the \emph{conflicting property} (P2306) \emph{instance of} (P31) with the \emph{conflicting value} (P2305) \emph{Wikimedia disambiguation page} (Q4167410). It is set because editors might easily confuse a work of art with a specific title, e.g., \emph{Tempest} (Q86440) by \emph{author} \emph{William Shakespear} (Q692) with the Wikimedia disambiguation page \emph{Tempest} (Q1589529).
\end{example}

\paragraph{Rules}
The conflicts-with constraint is violated if another statement on the same item either has a conflicting property or a conflicting property and a conflicting value depending on specified qualifiers. It depends on a conjunction (\(O\) denoting another statement ID, \(C\) another value) and a violation for a conflicts-with constraint is thus either rule one for constraints with no conflicting values specified or one instance of rule two for each conflicting value if specified.
\begin{align}
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow\\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{}\\
&\text{statement\_EDB}(O, E, \text{conflictingPropertyConstant}, C)
\end{split}\\
\begin{split}
\text{violation\_statement}&(S, E, \text{propertyConstant}, V) \leftarrow\\
&\text{statement\_EDB}(S, E, \text{propertyConstant}, V) \wedge{}\\
&\text{statement\_EDB}(O, E, \text{conflictingPropertyConstant}, \text{conflictingValueConstant})
\end{split}
\end{align}

\subsection{Allowed entity types (Q52004125)}
\paragraph{Constraint}
The \emph{allowed entity types} constraint limits if a property may be used in a statement whose subject is an item, a property, or a lexeme, although as of September 2018 lexemes are not deployed.

>>>>

\chapter{Preliminaries}\label{cha_preliminaries}
\section{Wikidata}\label{sec_wikidata}
Wikidata is the knowledge base sister project of Wikipedia. It is a public, open database system that, at its core, stores statements about specific items. The basic unit of information is a triple subject-predicate-object stating an item (subject) has a property (predicate) which is either another item or a value of some kind (object). Triples store the relations of entities with each other or with specific data. An entity is either an item, which represents a topic, concept or real world entity, or a property, which denotes a relation an item has to another or with a value. They have numeric identifiers that are prefixed with Q and P and will be referred to as Q- and P-IDs, respectively. In the future there will be a third entity type, lexeme. This is already incorporated into some constraints, but as of September 2018 they are not deployed. A value is one of several data types shown in Table \ref{tab_datatypes}, although it is also possible to use \textsc{none} or \textsc{some} as values for statements that have no value (akin to negation) or have a currently unknown value. Every property determines the data type it accepts as an object. \\

@@

While this is similar to RDF and Wikidata has in fact been exported as RDF \cite{EGKMV2014}, the data structure is more complex than simple triples. It allows to store additional information about statements. This is done via a statement ID, which is assigned to every triple storing information about an item. Statements have one of three ranks, \textsc{normal}, \textsc{preferred}, and \textsc{deprecated} which are used to differentiate similar statements on the same item. For example, the \textsc{preferred} rank would be given to the current \textsc{head of state} (P35) of a country, while all statements denoting former \textsc{heads of state} have the rank \textsc{normal}. \textsc{Deprecated} is used on statements that where once considered correct but are not anymore, e.g., \textsc{Chelsea Manning} (Q298423) having the \textsc{sex or gender} (P21) \textsc{male} (Q6581097).

Every statement can have added qualifiers, which consist of the corresponding statement id, a predicate, and an object. Qualifiers provide contextual information for statements. The most common context is the time frame in which the statement was accurate, e.g., stating the \textsc{start time} (P580) and \textsc{end time} (P582) of the term of a head of state, but other information is possible, e.g. determination method for measurements or the elevation above sea level for the highest point. Additionally, references can be added to statements.

@@

\begin{example}
Figure \ref{fig_example} is an excerpt from the entity page of the clipper \textsc{Cutty Sark} (Q171255) \footnote{\url{https://www.wikidata.org/wiki/Q171255}}. An entity page lists all statements with the entity as the subject. It then groups all statements by their property, in this case \textsc{length} (P2043). For the length of the Cutty Sark there is only one value, 85.4 metres, but this could include multiple values, e.g., if there were different ways to determine the length. This statement is further qualified with \textsc{criterion used} (P1013) as the qualifier property, which specifies how the value was determined. In this case the qualifier value \textsc{length overall} (Q2358152) was taken. The statement is backed up by one reference citing the property \textsc{reference URL} (P854) pointing to the reference value, which is an URL confirming the length of 85.4 metres.

@@

\section{Datalog}
Datalog is a logic query language. It is fully declarative and uses rules to derive additional knowledge from given facts. The most basic domain element in datalog is a \textsc{constant}, which will be denoted as \textsc{someConstant}. In the context of Wikidata, any item or property, e.g. \textsc{Earth} (Q2) or \textsc{head of government} (P6), could be represented as a constant. Relations in datalog are expressed using \textsc{predicates} with a specified arity \(n\), written as predicate\([n]\). To stay close to Wikidata in our examples, let us take the predicate triple\([3]\) as an example. This would be used to express the \textsc{item}-\textsc{property}-\textsc{value} relation with each position representing one part of the triple.\\
Additionally, datalog also provides \textsc{variables}, which will be capitalized, like \(X\). The sets of predicates {\bf P}, constants {\bf C} and variables {\bf V} are mutually disjoint. Constants and variables are collectively known as \textsc{terms} and are, together with predicates, used to form \textsc{atoms}. An atom has the form predicate\((t_1, \ldots , t_n)\) with \(n\) as the arity of the predicate and \(t_1,\ldots,t_n\) terms.\\
Atoms are the building blocks of \textsc{rules}, which are used to derive information in datalog. A rule is a premise that implies a conclusion, and both are conjunctions of atoms.
Let \(P_n\) denote  premise atoms and \(C_n\) the conclusion atoms. The variables used in one rule are divided into two mutually disjoint sets \(V_p\) and \(V_r\) which are the variables occurring only in its premise atoms and occurring in conclusion atoms and premise atoms, respectively. The general from of a rule is (with \(k \in \mathbb{N}^+\), \(l \in \mathbb{N}\))
\begin{equation}
\forall V_p \forall V_r.(C_1 \wedge \ldots \wedge C_k \leftarrow P_1 \wedge \ldots \wedge P_l)
\end{equation}

If \(V_c = \emptyset\) the rule is called a \textsc{datalog rule} and if \(l = 0\) it is called a \textsc{fact}. If a fact contains no variables but only constants, it is a \textsc{ground rule}. Since all \(V_p\) and \(V_r\) are  quantified, the preceding universal quantifiers \(\forall V_p \forall V_r\) will be omitted for the rules presented in this work.

Within the rule predicates we can differentiate between two different schemas: the \textsc{extensional schema} and the \textsc{intensional schema}, called \textsc{edb} and \textsc{idb} (extensional/intensional database). An \textsc{edb} predicate may only occur in the premise of a rule, meaning it can never be derived. All predicates that appear in the conclusion at least once are \textsc{idb} predicates.

\begin{example}
For the following example, taken from Wikidata, descriptive names will be used instead of the IDs to increase readability. The individual entities and properties are introduced with their ID beforehand, so the example can be retraced. The species \textsc{blue whale} (Q42196) has the \textsc{parent taxon} (P171) \textsc{balaenoptera} (Q133320), which in turn has the parent taxon \textsc{rorqual} (Q232829). The predicates used are triple\_edb\([3]\) as the input and isInTaxon\_idb\([2]\), which will be derived. The basic facts are:
\[\textsc{triple\_edb}(\textsc{blue whale}, \textsc{parent taxon}, \textsc{balaenoptera})\]
\[\textsc{triple\_edb}(\textsc{balaenoptera}, \textsc{parent taxon}, \textsc{rorqual})\]
From this we can draw conclusions. The first step is to represent the already existing facts as idb:
\[\textsc{isInTaxon\_idb}(X, Y) \leftarrow \textsc{triple\_edb}(X, \textsc{parent taxon}, Y)\]
Finally the transitive closure of this relation can be derived with
\[\textsc{isInTaxon\_idb}(X, Y) \leftarrow \textsc{isInTaxon\_idb}(X, Z) \wedge \textsc{isInTaxon\_idb}(Z, Y)\]
\end{example}

@@

To apply a constraint to a property, a statement <property> <P2302> <constraint> is added to the properties item page, where <property> is the property that should be constrained, <P2302> (property constraint) denotes that the property has a constraint and <constraint> is the specific constraint. Note that a constrained statement means a statement whose predicate is a property that has a specific constraint. A simple example for this would be \textsc{distinct values} (Q21502410), stating that values of the property must be unique across all constrained statements. For this case no additional information regarding the constraint is necessary and violation could be found by surveying all constrained statements and comparing their values.\\
Constraints can be further specified by using qualifiers on the constraint statement. These qualifiers are either optional or mandatory, and may be applied once or multiple times. For example, the \textsc{none of} constraint (Q52558054) states that values for the specific property may not be in a given set of disallowed values. This set is specified using qualifiers with property \textsc{item of} (P2305), which list all the disallowed values. At least one such qualifier must be present for the statement.\\
More complex dependencies are also possible. A constraint can depend on all constrained statements of a specific item (\textsc{single value} constraint [Q19474404]), on all statements of the item specified as the value of a constrained statement (\textsc{value requires statement} constraint [Q21510864]), or on all statements using the subclass of-predicate (\textsc{type} constraint [Q21503250]).

@@

\section{Rules}
Table \ref{tab_property_constraints} shows that most constraints cannot be modelled without at least some amount of preprocessing. Only \textsc{scope}, \textsc{conflicts with}, \textsc{allowed entity types} and \textsc{none of} can be modelled directly. This is due to the fact that \textsc{conflicts with} and \textsc{none of} exclude a limited and specified set of statements from validity and these cases can thus be directly taken as violations. \textsc{scope} and \textsc{allowed entity types} work very similar, since there are only three respectively two possible values and violations are simply the options that have not been allowed. The rules for these constraints are described in Section \ref{sec_simple_constraints}.\\

A problem that occurs multiple times is the need to determine inequality between constants. \textsc{distinct value}, \textsc{allowed units}, \textsc{allowed qualifiers}, \textsc{one of}, \textsc{single best value} and \textsc{one-of qualifier value} each require a rule to determine if a value, unit, or property is not equal to another. To this end a certain amount of preprocessing is necessary, most notably the extraction of all constant pairs whose inequality has to be determined for finding the violations. To avoid putting and undue amount of processing outside of datalog, this extraction will not check more than that the constant is in the right position relative to the constrained statement, e.g., all values on statements constrained by \textsc{distinct value} or all qualifiers on statements constrained by \textsc{allowed qualifiers}. From this set inequality will be generated, different versions of which will be introduced in Section \ref{sec_inequality} and evaluated with regards to their performance in Section \ref{sec_constraints_using_inequality}.\\

Negation as a general problem cannot be solved in datalog. One could pre-process every necessary fact to be negated and add this to the ground facts, but this is not the purpose of this work. However, not all constraints require a general negation. \textsc{item requires statement}, \textsc{value requires statement}, \textsc{inverse }and \textsc{symmetric} are all violated if a statement does not exist on a specific item. This problem can be solved with limited amounts of preprocessing by leveraging a preprocessed statement order per item and 'iterating' through the statements until the required conditions are met.\\
A similar approach can be taken to find items with exactly one statement matching certain criteria, required for 	\textsc{multi-value}. The idea is to ask the question if a second statement exists, then evaluating it as above. Both methods will be explained in Section \ref{sec_constraints_using_statement_non}.

\subsection{Representation of Wikidata as Facts}\label{subsec_representation}
To represent the content of Wikidata as facts, the following predicates are introduced:
\begin{table}[H]
\caption{\textsc{edb} predicates}\label{tab_predicates}
\begin{tabular}{llp{8.5cm}}
Predicate & Positions & Description \\
\hline
\textsc{statement\_edb} & \makecell{statementID \\ entityID \\ propertyID \\ value (entityID or string)} & represents basic statements \\
\hline
\textsc{qualifier\_edb} & \makecell{statementID \\ propertyID \\ value} & represents qualifiers on statements \\
\hline
\textsc{reference\_edb} & \makecell{statementID \\ reference hash \\ propertyID \\ value} & represents references on statements; reference hash groups the property-value pairs belonging to one reference \\
\hline
\textsc{item\_edb} & entityID & marks an entity as an item \\
\textsc{property\_edb} & entityID & marks an entity as a property \\
\hline
\textsc{unit\_edb} & \makecell{value \\ unitID} & denotes which unit a value has \\
\hline
\textsc{rank\_edb} & \makecell{statementID \\ rank} & denotes the rank a statement has
\end{tabular}
\end{table}

@@

\subsection{Task}\label{subsec_task}
Each property constraint will be modelled and evaluated as one set of rules per property per constraint, meaning each violating statement will be of a form similar to
\begin{align*}
&\textsc{violation\_statement}(S, E, \textsc{propertyConstant}, V) &\leftarrow \textsc{atoms determining the violation} \\
&\textsc{violation\_qualifier}(S, \textsc{propertyConstant}, V) &\leftarrow \textsc{atoms determining the violation} \\
&\textsc{violation\_reference}(S, H, \textsc{propertyConstant}, V) &\leftarrow \textsc{atoms determining the violation}
\end{align*}
where propertyConstant denotes the constrained property. The variables \(S\), \(I\), \(V\), and \(H\) are used for statement ID, entity ID, value and reference hash, respectively.

The task for each constraint is thus: Given the constraint and a constrained property, potentially with qualifiers, which rule finds all violations of the rule in a given set of triples, qualifiers and references if applicable? The given rule set should return all violations if queried with the following atoms:
\begin{align*}
&\textsc{violation\_statement}(S, E, P, V) \\
&\textsc{violation\_qualifier}(S, P, V) \\
&\textsc{violation\_reference}(S, H, P, V)
\end{align*}

@@

{\bf Constraint:} This subsection starts with a table showing basic properties of the constraint. These include the required data, relevant qualifiers, and possible violations of the constraint. The required data states which part of the data in relation to the constrained statement are restricted or need to exist. The qualifiers list their cardinality, meaning if they are required and if they may be used multiple times, and a short description. The violations list the conflicting (part of a) statement in relation to the constrained statement. The first constraint, \textsc{property scope}, will be used as an example to explain the table in detail.

@@

\subsection{Property scope (Q53869507)}
\paragraph{Constraint}
The \textsc{property scope} constraint specifies if a property may be used as predicate in a statement, qualifier, or reference.
\begin{table}[H]
\caption{property scope constraint (Q53869507)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Required data & constrained statement, qualifier and reference \\
\hline
Qualifiers & \makecell{\textsc{property scope} (P5314) -- 1..3 \\ must be \textsc{as main value} (Q54828448), \textsc{as qualifier} (Q54828449), \\ or \textsc{as reference} (Q54828450) \\ specifies allowed use in statements, qualifiers, or references} \\
\hline
Violation & constrained property in a position that is not allowed \\
\hline
\end{tabularx}
\end{table}

The required data specifies which data, in relation to the constrained statement, needs to be evaluated to conclude if the constraint was violated. The used abbreviations are listed in Table \ref{tab_abbreviations}. In this case, this is simply each statement, qualifier, or reference whose predicate is the constrained property.\\
The \textsc{property scope} constraint requires the scope, meaning statement, qualifier, or reference, to be set via qualifier \textsc{property scope} (P5314). The cardinality is 1..3 because it is mandatory and has only three possible values.\\
It is violated if a statement, qualifier, or reference has a predicate with the \textsc{property scope} constraint that does not allow this scope.\\

\begin{example}
The \textsc{reference URL} (P854) property is restricted to the \textsc{property scope} (P5314) \textsc{as reference} (Q54828450), because it is to be used to mark a URL that substantiates a statement.
\end{example}

\paragraph{Rules}
The \textsc{property scope} constraint requires no context information to be violated; the triple, qualifier, or reference with the constrained property is enough. This means violations can be found using these rules:
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \textsc{propertyConstant}, V) \leftarrow \\
&\textsc{statement\_edb}(S, E, \textsc{propertyConstant}, V)\\
\end{split}\\
\begin{split}
\textsc{violation\_qualifier}&(S, \textsc{propertyConstant}, V) \leftarrow \\
&\textsc{qualifier\_edb}(S, \textsc{propertyConstant}, V)\\
\end{split}\\
\begin{split}
\textsc{violation\_reference}&(S, \textsc{propertyConstant}, V) \leftarrow \\
&\textsc{reference\_edb}(S, H, \textsc{propertyConstant}, V)\\
\end{split}
\end{align}
Which of these rules need to be used depends on the values of the \textsc{property scope} qualifier. If a concrete constrained on a property does not have \textsc{as main value} as a value of the qualifier, the first rule is used. The case for \textsc{as qualifier} and \textsc{as reference} is analogous with the second and third rule, respectively.

\subsection{Conflicts with (Q21502838)}
\paragraph{Constraint}
The \textsc{conflicts with} constraint forbids statements on constrained entities whose predicate is a specified \textsc{conflicting property}. If \textsc{conflicting values} have been specified as well, the constraint is weakened and only violated by statements if in addition to the predicate being a \textsc{conflicting property}, the object is one of the \textsc{conflicting values}.
\begin{table}[H]
\caption{Conflicts-with constraint (Q21502838)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Required data & entity statements \\
\hline
Qualifiers & \makecell{\textsc{conflicting property} (P2306) -- 1 \\ property that may not be predicate of entity statements \\
\textsc{conflicting values} (P2305) -- 0..* \\ values that may not be objects of entity statements with the \textsc{conflicting property}} \\
\hline
Violation & \makecell{entity statement with the conflicting property \\ entity statement with the conflicting property and one of the conflicting values} \\
\hline
\end{tabularx}
\end{table}

\begin{example}
The \textsc{author} (P50) constraint has (among others) the \textsc{conflicting property} (P2306) \textsc{instance of} (P31) with the \textsc{conflicting value} (P2305) \textsc{Wikimedia disambiguation page} (Q4167410). It is set because editors might easily confuse a work of art with a specific title, e.g., \textsc{Tempest} (Q86440) by \textsc{author} \textsc{William Shakespear} (Q692) with the Wikimedia disambiguation page \textsc{Tempest} (Q1589529).
\end{example}

\paragraph{Rules}
The conflicts-with constraint is violated if another statement on the same item either has a conflicting property or a conflicting property and a conflicting value depending on specified qualifiers. It depends on a conjunction (\(O\) denoting another statement ID, \(C\) another value) and a violation for a conflicts-with constraint is thus either rule one for constraints with no conflicting values specified or one instance of rule two for each conflicting value if specified.
\begin{align}
\begin{split}
\textsc{violation\_statement}&(S, E, \textsc{propertyConstant}, V) \leftarrow\\
&\textsc{statement\_edb}(S, E, \textsc{propertyConstant}, V) \wedge{}\\
&\textsc{statement\_edb}(O, E, \textsc{conflictingPropertyConstant}, C)
\end{split}\\
\begin{split}
\textsc{violation\_statement}&(S, E, \textsc{propertyConstant}, V) \leftarrow\\
&\textsc{statement\_edb}(S, E, \textsc{propertyConstant}, V) \wedge{}\\
&\textsc{statement\_edb}(O, E, \textsc{conflictingPropertyConstant}, \textsc{conflictingValueConstant})
\end{split}
\end{align}

\subsection{Allowed entity types (Q52004125)}
\paragraph{Constraint}
The \textsc{allowed entity types} constraint limits if a property may be used in a statement whose subject is an item, a property, or a lexeme, although as of September 2018 lexemes are not deployed.

\begin{table}[H]
\caption{Allowed entity types constraint (Q52004125)}
\begin{tabularx}{\textwidth}{ ll X}
\hline
Required data & constrained statements and entity types\\
\hline
Qualifiers & \makecell{allowed entity type (P2305) -- 1..3 \\ must be \textsc{Wikibase item} (Q29934200), \textsc{Wikibase property} (Q29934218), \\ or lexeme (Q51885771); \\ specifies the entity type the property may be used on} \\
\hline
Violation & \makecell{constrained statement on an entity of a type unequal to all allowed types} \\
\hline
\end{tabularx}
\end{table}

Since there are only three options (and only two in use), violations can be found without inequality by finding the \textsc{statement\_edb} with the constrained property whose subject is of the type that is not allowed, meaning either \textsc{Wikibase item} or \textsc{Wikibase property}.
